---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RcppColMetric

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/RcppColMetric)](https://CRAN.R-project.org/package=RcppColMetric)
[![R-CMD-check](https://github.com/zhuxr11/RcppColMetric/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zhuxr11/RcppColMetric/actions/workflows/R-CMD-check.yaml)
[![Download stats](https://cranlogs.r-pkg.org/badges/grand-total/RcppColMetric)](https://CRAN.R-project.org/package=RcppColMetric)
<!-- badges: end -->

**Package**: [*RcppColMetric*](https://github.com/zhuxr11/RcppColMetric) `r pkgload::pkg_version()`<br />
**Author**: Xiurui Zhu<br />
**Modified**: `r file.info("README.Rmd")$mtime`<br />
**Compiled**: `r Sys.time()`

The goal of `RcppColMetric` is to efficiently compute metrics between various vectors and a common vector. This is common in data science, such as computing performance metrics between each feature and a common response. [`Rcpp`](https://CRAN.R-project.org/package=Rcpp) is used to efficiently iterate over vectors through compiled code. You may extend its utilities by providing custom metrics that fit into the framework.

## Installation

You can install the released version of `RcppColMetric` from [CRAN](https://cran.r-project.org/) with:

``` r
install.packages("RcppColMetric")
```

Alternatively, you can install the developmental version of `RcppColMetric` from [github](https://github.com/) with:

``` r
remotes::install_github("zhuxr11/RcppColMetric")
```

## Examples

```{r library, eval=FALSE}
library(cbbinom)
```

```{r, include=FALSE}
pkgload::load_all()
```

We use `cats` from [`MASS`](https://CRAN.R-project.org/package=MASS) to illustrate the use of the package.

```{r cats-dataset}
library(MASS)
data(cats)
print(head(cats))
```

### Column-wise ROC-AUC

In binary classification modelling, it is a common practice to compute ROC-AUC of each feature (usually columns) against a common target. `RcppColMetric` provides a much faster version than its commonly used counterparts, e.g. `caTools::colAUC()`.

```{r col-auc}
library(caTools)
(col_auc_bench <- microbenchmark::microbenchmark(
  col_auc_r = caTools::colAUC(cats[, 2L:3L], cats[, 1L]),
  col_auc_cpp = col_auc(cats[, 2L:3L], cats[, 1L]),
  times = 100L,
  check = "identical"
))
```

As can be seen, the median speed of computation from `RcppColMetric` is `r round(summary(col_auc_bench)[["median"]][1L] / summary(col_auc_bench)[["median"]][2L], 3L)` times faster.

If there are multiple sets of features and responses, you may use the vectorized version `col_auc_vec()`, which uses compiled code to speed up iterations and returns a list.

```{r col-auc-vec}
col_auc_vec(list(cats[, 2L:3L]), list(cats[, 1L]))
```

### Column-wize mutual information

In classification modelling, it is another common practice to assess mutual information between features and a response if the features are discrete. `RcppColMetric` provides a much faster version than its commonly used counterparts, e.g. `infotheo::mutinformation()`.

```{r col-mut-info}
library(infotheo)
(col_mut_info_bench <- microbenchmark::microbenchmark(
  col_mut_info_r = sapply(round(cats[, 2L:3L]), infotheo::mutinformation, cats[, 1L]) %>%
    {matrix(., nrow = 1L, dimnames = list(NULL, names(.)))},
  col_mut_info_cpp = col_mut_info(round(cats[, 2L:3L]), cats[, 1L]),
  times = 100L,
  check = "identical"
))
```

As can be seen, the median speed of computation from `RcppColMetric` is `r round(summary(col_mut_info_bench)[["median"]][1L] / summary(col_mut_info_bench)[["median"]][2L], 3L)` times faster.

If there are multiple sets of features and responses, you may use the vectorized version `col_mut_info_vec()`, which uses compiled code to speed up iterations and returns a list.

```{r col-mut-info-vec}
col_mut_info_vec(list(round(cats[, 2L:3L])), list(cats[, 1L]))
```

## Extend the package with custom metric

You may implement your own metric by inheriting from `RcppColMetric::Metric` class with template arguments as feature `SEXP` (input, numeric here) and response `SEXP` (input, factor as integer here) types. For example, to compute range of each feature, define a `RangeMetric` class.

```{Rcpp, include=FALSE}
#include "E:/R/Packages/RcppColMetric/inst/include/RcppColMetric.h"
#include <Rcpp.h>
using namespace Rcpp;

class 
RangeMetric: public RcppColMetric::Metric<REALSXP, INTSXP>
{
public:
  // Constructor
  RangeMetric(const RObject& x, const IntegerVector& y, const Nullable<List>& args = R_NilValue) {
  // This parameter is inherited from `Metric`, determining output dimension (number of rows)
  // For RangeMetric, the output dimension is 1 (scalar)
  output_dim = 1;
}
  
  virtual Nullable<CharacterVector> row_names(const RObject& x, const IntegerVector& y, const Nullable<List>& args = R_NilValue) const override {
    // Determine the row names
    // For RangeMetric, the row names are NULL (or this can be CharacterVector)
    return R_NilValue;
  }
  virtual NumericVector calc_col(const NumericVector& x, const IntegerVector& y, const R_xlen_t& i, const Nullable<List>& args = R_NilValue) const override {
    // Derive output value for each feature and the common response
    // For RangeMetric, the output is max - min
    NumericVector out = {max(x) - min(x)};
    return out;
  }
};

// [[Rcpp::export]]
NumericMatrix col_range(const RObject& x, const IntegerVector& y, const Nullable<List>& args = R_NilValue) {
  RangeMetric range_metric(x, y, args);
  NumericMatrix out = RcppColMetric::col_metric<REALSXP, INTSXP, REALSXP>(x, y, range_metric, args);
  return out;
}

RangeMetric gen_range_metric(const RObject& x, const IntegerVector& y, const Nullable<List>& args = R_NilValue) {
  RangeMetric out(x, y, args);
  return out;
}

// [[Rcpp::export]]
List col_range_vec(const List& x, const List& y, const Nullable<List>& args = R_NilValue) {
  List out = RcppColMetric::col_metric_vec<REALSXP, INTSXP, REALSXP>(x, y, &gen_range_metric, args);
  return out;
}
```

```{Rcpp range-metric, eval=FALSE}
#include <RcppColMetric.h>
#include <Rcpp.h>
using namespace Rcpp;

class 
RangeMetric: public RcppColMetric::Metric<REALSXP, INTSXP>
{
public:
  // Constructor
  RangeMetric(const RObject& x, const IntegerVector& y, const Nullable<List>& args = R_NilValue) {
  // This parameter is inherited from `Metric`, determining output dimension (number of rows)
  // For RangeMetric, the output dimension is 1 (scalar)
  output_dim = 1;
}
  
  virtual Nullable<CharacterVector> row_names(const RObject& x, const IntegerVector& y, const Nullable<List>& args = R_NilValue) const override {
    // Determine the row names
    // For RangeMetric, the row names are NULL (or this can be CharacterVector)
    return R_NilValue;
  }
  virtual NumericVector calc_col(const NumericVector& x, const IntegerVector& y, const R_xlen_t& i, const Nullable<List>& args = R_NilValue) const override {
    // Derive output value for each feature and the common response
    // For RangeMetric, the output is max - min
    NumericVector out = {max(x) - min(x)};
    return out;
  }
};
```

Then, define the main function calling `RcppColMetric::col_metric()`, with corresponding feature `SEXP` (input, numeric here), response `SEXP` (input, factor as integer here) and output `SEXP` (output, numeric here) types.

```{Rcpp col-range, eval=FALSE}
NumericMatrix col_range(const RObject& x, const IntegerVector& y, const Nullable<List>& args = R_NilValue) {
  RangeMetric range_metric(x, y, args);
  NumericMatrix out = RcppColMetric::col_metric<REALSXP, INTSXP, REALSXP>(x, y, range_metric, args);
  return out;
}
```

Test this function with `cats`:

```{r col-range-example}
col_range(cats[, 2L:3L], cats[, 1L])
```

To define vectorized version of the function, a wrapper function is defined to generate `RangeMetric` object, and then passed on to the workhorse `RcppColMetric::col_metric_vec()`.

```{Rcpp col-range-vec, eval=FALSE}
RangeMetric gen_range_metric(const RObject& x, const IntegerVector& y, const Nullable<List>& args = R_NilValue) {
  RangeMetric out(x, y, args);
  return out;
}

// [[Rcpp::export]]
List col_range_vec(const List& x, const List& y, const Nullable<List>& args = R_NilValue) {
  List out = RcppColMetric::col_metric_vec<REALSXP, INTSXP, REALSXP>(x, y, &gen_range_metric, args);
  return out;
}
```

Test the vectorized function with `cats`:

```{r col-range-vec-example}
col_range_vec(list(cats[, 2L:3L]), list(cats[, 1L]))
```

